\documentclass{beamer}

\usetheme{Madrid}
\usepackage{amsmath,amssymb,physics}

\title{Entanglement-Enhanced Quantum Sensing}
\subtitle{Singlet--Triplet Interferometry, QFI, and Decoherence}
\author{MHJ}
\date{}

\begin{document}

%-------------------------------------------------
\begin{frame}
\titlepage
\end{frame}

%=================================================
\section{Physical mechanism}

\begin{frame}{Entangled probe state}
We start from the spin singlet state
\[
\ket{S} = \frac{1}{\sqrt{2}}
\left(
\ket{\uparrow\downarrow} - \ket{\downarrow\uparrow}
\right),
\]
which has total spin zero and is invariant under common-mode magnetic
fields.

\vspace{0.3cm}

This makes the singlet a \textbf{noise-protected reference state} for
differential sensing.
\end{frame}

%-------------------------------------------------
\begin{frame}{Phase accumulation by a local magnetic field}
A magnetic field acting only on electron 1 induces a relative phase
\[
\ket{\uparrow}_1 \rightarrow e^{+i\phi/2}\ket{\uparrow}_1,
\qquad
\ket{\downarrow}_1 \rightarrow e^{-i\phi/2}\ket{\downarrow}_1,
\]
with
\[
\phi(t) = \gamma \int_0^t B(t')\,dt'.
\]

Applying this to the singlet gives
\[
\ket{\psi(\phi)} =
\frac{1}{\sqrt{2}}
\left(
e^{i\phi/2}\ket{\uparrow\downarrow}
-
e^{-i\phi/2}\ket{\downarrow\uparrow}
\right).
\]
\end{frame}

%-------------------------------------------------
\begin{frame}{Singlet--triplet decomposition}
Rewrite $\ket{\psi(\phi)}$ in the singlet--triplet basis:
\[
\ket{\psi(\phi)} =
\cos\!\left(\frac{\phi}{2}\right)\ket{S}
+
i\sin\!\left(\frac{\phi}{2}\right)\ket{T_0},
\]
where
\[
\ket{T_0} = \frac{1}{\sqrt{2}}
\left(
\ket{\uparrow\downarrow} + \ket{\downarrow\uparrow}
\right).
\]

\vspace{0.3cm}

\textbf{Key point:} the phase $\phi$ is converted into a measurable
population transfer.
\end{frame}

%=================================================
\section{Connection to Ramsey interferometry}

\begin{frame}{Relation to Ramsey interferometry}
This protocol is mathematically equivalent to a Ramsey experiment:
\begin{itemize}
  \item Singlet preparation $\leftrightarrow$ first $\pi/2$ pulse
  \item Phase accumulation under $B(t)$ $\leftrightarrow$ free evolution
  \item Singlet--triplet readout $\leftrightarrow$ final $\pi/2$ pulse
\end{itemize}

\vspace{0.3cm}

Unlike single-spin Ramsey interferometry:
\begin{itemize}
  \item common-mode noise cancels automatically,
  \item the interferometer operates in an \textbf{entangled basis}.
\end{itemize}
\end{frame}

%=================================================
\section{Quantum Fisher information}

\begin{frame}{Quantum Fisher information: definition}
For a pure state $\ket{\psi(\phi)}$, the quantum Fisher information is
\[
F_Q
=
4\left(
\braket{\partial_\phi\psi|\partial_\phi\psi}
-
\left|\braket{\psi|\partial_\phi\psi}\right|^2
\right).
\]

It sets the ultimate precision bound via the quantum Cramér--Rao inequality:
\[
\Delta \phi \ge \frac{1}{\sqrt{F_Q\,N}}.
\]
\end{frame}

%-------------------------------------------------
\begin{frame}{QFI for singlet--triplet interferometry}
Using
\[
\ket{\psi(\phi)} =
\cos\!\left(\frac{\phi}{2}\right)\ket{S}
+
i\sin\!\left(\frac{\phi}{2}\right)\ket{T_0},
\]
we compute
\[
\partial_\phi \ket{\psi}
=
-\frac{1}{2}\sin\!\left(\frac{\phi}{2}\right)\ket{S}
+
\frac{i}{2}\cos\!\left(\frac{\phi}{2}\right)\ket{T_0}.
\]

This yields
\[
F_Q = 1
\quad \text{(per interrogation time unit)}.
\]

With optimized control and extended interaction time, $F_Q$ scales with
the effective generator variance, leading to large enhancements.
\end{frame}

%-------------------------------------------------
\begin{frame}{Physical meaning of QFI here}
\begin{itemize}
  \item The generator of $\phi$ acts \emph{differentially} on the two spins.
  \item The singlet has maximal susceptibility to such differential phases.
  \item Entanglement amplifies phase sensitivity beyond separable probes.
\end{itemize}

\vspace{0.3cm}

This is the origin of the observed quantum metrological advantage.
\end{frame}

%=================================================
\section{Decoherence and triplet leakage}

\begin{frame}{Including decoherence}
Decoherence (e.g.\ dephasing) suppresses singlet--triplet coherence.
A simple phenomenological model gives
\[
\rho_{ST}(\phi,t)
=
\begin{pmatrix}
\cos^2(\phi/2) & i e^{-\Gamma t}\cos(\phi/2)\sin(\phi/2) \\
- i e^{-\Gamma t}\cos(\phi/2)\sin(\phi/2) & \sin^2(\phi/2)
\end{pmatrix}.
\]

Here $\Gamma$ is the decoherence rate.
\end{frame}

%-------------------------------------------------
\begin{frame}{Triplet leakage with decoherence}
The measured triplet population becomes
\[
P_{T_0}(\phi,t)
=
\frac{1}{2}
\left(
1 - e^{-\Gamma t}\cos\phi
\right).
\]

For small $\phi$:
\[
P_{T_0} \approx
\frac{\phi^2}{4}\,e^{-\Gamma t}.
\]

\vspace{0.3cm}

Decoherence reduces contrast but preserves phase sensitivity.
\end{frame}

%-------------------------------------------------
\begin{frame}{Impact on quantum Fisher information}
Decoherence reduces the QFI to
\[
F_Q(t) = F_Q(0)\,e^{-2\Gamma t}.
\]

\begin{itemize}
  \item Short times: quantum advantage retained
  \item Long times: decoherence restores shot-noise scaling
\end{itemize}

This defines the optimal sensing time.
\end{frame}

%=================================================
\section{Summary}

\begin{frame}{Summary}
\begin{itemize}
  \item Local magnetic fields induce singlet--triplet mixing.
  \item Phase information is converted into measurable populations.
  \item The protocol is an entanglement-based Ramsey interferometer.
  \item Quantum Fisher information quantifies the enhanced sensitivity.
  \item Decoherence limits, but does not invalidate, the quantum advantage.
\end{itemize}
\end{frame}

\end{document}
