\documentclass[aspectratio=169]{beamer}
\usetheme{Madrid}
\usecolortheme{default}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}[frame number]

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,mathtools,bm,amsfonts}
\usepackage{booktabs}
\usepackage{array}
\usepackage{physics}
\usepackage{graphicx}
\usepackage{tikz}

\usetikzlibrary{arrows.meta,positioning}

\title{Reinforcement Learning for Quantum Control}
%\subtitle{A Physics-Oriented Introduction with Applications to Quantum Sensors}
\author{Morten Hjorth-Jensen}
\date{March 17, 2026}

\begin{document}

%================================================
\frame{\titlepage}

%================================================
\begin{frame}{Outline}
\begin{enumerate}
\item Why reinforcement learning?
\item Basic ingredients: states, actions, rewards, policies
\item Value functions and Bellman equations
\item Q-learning, policy gradients, actor--critic
\item Worked RL example with equations and pseudocode
\item Model-aware reinforcement learning
\item Reinforcement learning for quantum control
\item Applications to quantum sensing if time
\end{enumerate}
\end{frame}

%================================================
\begin{frame}{Motivation}
Many problems in physics can be phrased as optimization problems:
\begin{itemize}
\item variational calculations,
\item control of dynamical systems,
\item adaptive measurements,
\item maximizing sensitivity in quantum sensors.
\end{itemize}

However, in many realistic settings:
\begin{itemize}
\item the objective is not known analytically,
\item gradients are unavailable or noisy,
\item decisions must be taken sequentially,
\item present choices affect future opportunities.
\end{itemize}

Reinforcement learning (RL) addresses precisely this type of problem.
\end{frame}


\begin{frame}{References}
  \begin{itemize}
  \item R. Sutton and A. Barto, \textit{Reinforcement Learning: An Introduction}, \url{https://web.stanford.edu/class/psych209/Readings/SuttonBartoIPRLBook2ndEd.pdf}
  \item Marin Bukov and Florian Marquardt, \textit{Reinforcement Learning for Quantum Technology}, arXiv:2601.18953.
  \item  Federico Belliardo, Fabio Zoratti, Florian Marquardt, and Vittorio Giovannetti, \textit{Model-aware reinforcement learning for high-performance Bayesian experimental design in quantum metrology}, Quantum {\bf 8}, 1555 (2024).
  \end{itemize}

\end{frame}


\begin{frame}{My more narrow motivation (another talk perhaps)}
\begin{itemize}
  \item Quantum sensors permit fine-grained control over probe evolution and measurement settings.
  \item The central design problem is sequential: which control should be applied \emph{next}, given what has already been learned?
  \item For realistic platforms, the estimation stack contains many non-smooth components: sampling of measurement outcomes, particle-filter resampling, and optional weak-measurement backaction.
  \item The article by Belliardo {\em et al.}  proposes a \alert{model-aware RL} framework that treats experimental design as a trainable control problem while retaining a faithful physical model.
\end{itemize}
\end{frame}

\begin{frame}{Main thesis of the paper of Belliardo et al.}
\begin{block}{Core idea}
Use a differentiable simulation of the experiment, Bayesian inference by particle filtering, and policy optimization to learn adaptive or non-adaptive control laws for quantum metrology.
\end{block}
\begin{itemize}
  \item \textbf{Bayesian mode:} optimize the final estimation error after a complete measurement sequence.
  \item \textbf{Frequentist mode:} optimize the Cramer-Rao bound through the Fisher information matrix.
  \item \textbf{Model-aware:} gradients are propagated through a known sensor model rather than inferred only from trial-and-error interaction.
\end{itemize}
\end{frame}

\begin{frame}{The measurement loop as a control architecture}
\begin{enumerate}
  \item From the current posterior summary, the agent outputs the next control $x_{t+1}$.
  \item The probe evolves and is measured, producing a stochastic outcome $y_{t+1}$.
  \item The Bayesian posterior is updated and fed back to the agent.
\end{enumerate}
\vspace{0.5em}
Adaptive policies are written as
\[
  x_{t+1}=F_{\lambda}\!\big(P(\theta\mid x_t,y_t);\, y_t;\, R_t;\, t\big),
\]
while non-adaptive strategies suppress dependence on the posterior and outcomes:
\[
  x_{t+1}=F_{\lambda}(R_t,t).
\]
\begin{itemize}
  \item $\lambda$: trainable parameters of the policy (typically a neural network).
  \item $R_t$: resources consumed up to step $t$ (time, photons, signal amplitude, \ldots).
\end{itemize}
\end{frame}




%================================================
\begin{frame}{Back to basic RL: What is Reinforcement Learning?}
  An agent interacts with an environment over discrete time steps:
  \[
  s_0,a_0,r_1,s_1,a_1,r_2,\dots
  \]

  Core ingredients:
  \begin{itemize}
  \item \textbf{State} $s_t$: information available at time $t$,
  \item \textbf{Action} $a_t$: decision taken by the agent,
  \item \textbf{Reward} $r_{t+1}$: feedback from the environment,
  \item \textbf{Policy} $\pi(a|s)$: rule for choosing actions.
  \end{itemize}

  Goal:
  \[
  \max_\pi \; \mathbb{E}_\pi\!\left[\sum_{t=0}^{\infty}\gamma^t r_{t+1}\right]
  \]
  where $0\le \gamma <1$ is a discount factor.
\end{frame}

%================================================
\begin{frame}{Sequential Decision Making}
  In ordinary optimization we choose parameters once.

  In reinforcement learning we choose a \emph{sequence} of actions:
  \[
  a_0,a_1,a_2,\dots
  \]

  This matters because:
  \begin{itemize}
  \item actions change the future state,
  \item the future state changes which actions are useful later,
  \item rewards may be delayed.
  \end{itemize}

  Thus RL is best viewed as \textbf{optimization over trajectories or strategies}.
\end{frame}

%================================================
\begin{frame}{Markov Decision Process}
  The standard framework is a Markov Decision Process (MDP).

  Elements:
  \begin{itemize}
  \item state space $\mathcal S$,
  \item action space $\mathcal A$,
  \item transition probability $P(s'|s,a)$,
  \item reward function $R(s,a)$ or $R(s,a,s')$.
  \end{itemize}

  Markov property:
  \[
  P(s_{t+1}|s_t,a_t,s_{t-1},a_{t-1},\dots)=P(s_{t+1}|s_t,a_t)
  \]

  This is analogous to stochastic evolution equations in statistical physics.
\end{frame}

%================================================
\begin{frame}{Return and Value Functions}
  Define the return from time $t$:
  \[
  G_t = \sum_{k=0}^\infty \gamma^k r_{t+k+1}
  \]

  The \textbf{state-value function} under policy $\pi$ is
  \[
  V^\pi(s)=\mathbb{E}_\pi[G_t\mid s_t=s]
  \]

  The \textbf{action-value function} is
  \[
  Q^\pi(s,a)=\mathbb{E}_\pi[G_t\mid s_t=s,\; a_t=a]
  \]

  The \textbf{advantage} is
  \[
  A^\pi(s,a)=Q^\pi(s,a)-V^\pi(s)
  \]

  Interpretation: how much better is action $a$ than the typical action at state $s$?
\end{frame}

%================================================
\begin{frame}{Bellman Equation}
  The value function satisfies a recursive relation:
  \[
  V^\pi(s)=\sum_a \pi(a|s)\sum_{s'} P(s'|s,a)
  \left[r(s,a,s')+\gamma V^\pi(s')\right]
  \]

  This is the Bellman equation.

  Interpretation:
  \[
  \text{value} = \text{immediate reward} + \text{discounted future value}
  \]

  This is a self-consistency relation, reminiscent of mean-field equations and iterative schemes in physics.
\end{frame}

%================================================
\begin{frame}{Optimal Bellman Equation}
  Define the optimal value
  \[
  V^*(s)=\max_\pi V^\pi(s)
  \]
  and optimal action-value function
  \[
  Q^*(s,a)=\max_\pi Q^\pi(s,a).
  \]

  Then
  \[
  V^*(s)=\max_a \sum_{s'} P(s'|s,a)
  \left[r(s,a,s')+\gamma V^*(s')\right]
  \]

  and
  \[
  Q^*(s,a)=\sum_{s'} P(s'|s,a)
  \left[r(s,a,s')+\gamma \max_{a'}Q^*(s',a')\right].
  \]

  This is the basis for dynamic programming and Q-learning.
\end{frame}

%================================================
\begin{frame}{Value Iteration and Policy Iteration}
  \textbf{Value iteration:}
  \[
  V_{k+1}(s)=\max_a \sum_{s'} P(s'|s,a)\left[r+\gamma V_k(s')\right]
  \]

  \textbf{Policy iteration:}
  \begin{enumerate}
  \item Evaluate $V^\pi$ for current policy,
  \item Improve the policy using
    \[
    \pi_{\text{new}}(s)=\arg\max_a Q^\pi(s,a)
    \]
  \end{enumerate}

  These are exact if the model is known and the state-action spaces are manageable.
\end{frame}

%================================================
\begin{frame}{Q-Learning}
  When the model is unknown, we can learn directly from sampled transitions.

  Q-learning update:
  \[
  Q(s_t,a_t)\leftarrow Q(s_t,a_t) +
  \alpha\Big[r_{t+1}+\gamma \max_{a'}Q(s_{t+1},a')-Q(s_t,a_t)\Big]
  \]

  Here:
  \begin{itemize}
  \item $\alpha$ is the learning rate,
  \item the term in brackets is the temporal-difference error.
  \end{itemize}

  This is a stochastic approximation to the Bellman optimality equation.
\end{frame}

%================================================
\begin{frame}{Exploration vs Exploitation}
  A central issue in RL is the exploration--exploitation tradeoff.

  \begin{itemize}
  \item \textbf{Exploitation}: choose the best-known action,
  \item \textbf{Exploration}: try other actions to discover better strategies.
  \end{itemize}

  Common strategies:
  \begin{itemize}
  \item $\varepsilon$-greedy,
  \item softmax/Boltzmann exploration,
  \item entropy regularization.
  \end{itemize}

  This is analogous to balancing local minimization and global search in complex optimization landscapes.
\end{frame}

%================================================
\begin{frame}{Why Policy Gradient Methods?}
  Q-learning is natural for discrete action spaces.

  But in physics and control problems we often have:
  \begin{itemize}
  \item continuous control amplitudes,
  \item stochastic policies,
  \item large state spaces,
  \item differentiable parameterizations.
  \end{itemize}

  Then it is natural to parameterize the policy directly:
  \[
  \pi_\theta(a|s)
  \]
  and optimize the expected return
  \[
  J(\theta).
  \]
\end{frame}

%================================================
\begin{frame}{Policy Gradient Setup}
  Define
  \[
  J(\theta)=\mathbb{E}_{\tau\sim \pi_\theta}[R(\tau)]
  \]
  where $\tau=(s_0,a_0,s_1,a_1,\dots)$ is a trajectory and
  \[
  R(\tau)=\sum_{t=0}^\infty \gamma^t r_{t+1}.
  \]

  Trajectory probability:
  \[
  P_\theta(\tau)=\rho(s_0)\prod_{t=0}^\infty \pi_\theta(a_t|s_t)P(s_{t+1}|s_t,a_t)
  \]

  We want to compute
  \[
  \nabla_\theta J(\theta).
  \]
\end{frame}

%================================================
\begin{frame}{Policy Gradient Theorem: Step 1}
  Start with
  \[
  J(\theta)=\sum_\tau P_\theta(\tau)R(\tau).
  \]

  Differentiate:
  \[
  \nabla_\theta J(\theta)=\sum_\tau \nabla_\theta P_\theta(\tau)\,R(\tau).
  \]

  Use the log-derivative identity:
  \[
  \nabla_\theta P_\theta(\tau)=P_\theta(\tau)\nabla_\theta \log P_\theta(\tau).
  \]

  Hence
  \[
  \nabla_\theta J(\theta)=\sum_\tau P_\theta(\tau)\nabla_\theta \log P_\theta(\tau)\,R(\tau).
  \]

  Therefore
  \[
  \nabla_\theta J(\theta)=\mathbb{E}_{\tau\sim \pi_\theta}\left[\nabla_\theta \log P_\theta(\tau)\,R(\tau)\right].
  \]
\end{frame}

%================================================
\begin{frame}{Policy Gradient Theorem: Step 2}
  Now expand the log trajectory probability:
  \[
  \log P_\theta(\tau)=\log \rho(s_0)+
  \sum_{t=0}^\infty \log \pi_\theta(a_t|s_t)+
  \sum_{t=0}^\infty \log P(s_{t+1}|s_t,a_t).
  \]

  Only the policy depends on $\theta$, so
  \[
  \nabla_\theta \log P_\theta(\tau)=
  \sum_{t=0}^\infty \nabla_\theta \log \pi_\theta(a_t|s_t).
  \]

  Substituting,
  \[
  \nabla_\theta J(\theta)=
  \mathbb{E}_{\tau\sim \pi_\theta}
  \left[
    \left(
    \sum_{t=0}^\infty \nabla_\theta \log \pi_\theta(a_t|s_t)
    \right)
    R(\tau)
    \right].
  \]
\end{frame}

%================================================
\begin{frame}{Policy Gradient Theorem: Step 3}
  We can refine this by assigning to each time step only the future return from that point onward:
  \[
  G_t=\sum_{k=t}^\infty \gamma^{k-t}r_{k+1}.
  \]

  Then
  \[
  \nabla_\theta J(\theta)=
  \mathbb{E}_{\pi_\theta}
  \left[
    \sum_{t=0}^\infty \nabla_\theta \log \pi_\theta(a_t|s_t)\,G_t
    \right].
  \]

  This is the basis of the REINFORCE algorithm.

  A more compact form is
  \[
  \nabla_\theta J(\theta)=
  \mathbb{E}_{s\sim d^\pi,\,a\sim \pi_\theta}
  \left[
    \nabla_\theta \log \pi_\theta(a|s)\,Q^{\pi_\theta}(s,a)
    \right]
  \]
  where $d^\pi(s)$ is the discounted state visitation distribution.
\end{frame}

%================================================
\begin{frame}{Variance Reduction and Baselines}
  The estimator can have high variance.

  We may subtract a baseline $b(s)$ without changing the expectation:
  \[
  \nabla_\theta J(\theta)=
  \mathbb{E}\left[
    \nabla_\theta \log \pi_\theta(a|s)\left(Q^\pi(s,a)-b(s)\right)
    \right].
  \]

  A natural choice is
  \[
  b(s)=V^\pi(s),
  \]
  which gives
  \[
  \nabla_\theta J(\theta)=
  \mathbb{E}\left[
    \nabla_\theta \log \pi_\theta(a|s)\,A^\pi(s,a)
    \right].
  \]

  This motivates actor--critic methods.
\end{frame}

%================================================
\begin{frame}{REINFORCE Algorithm}
  \textbf{Basic Monte Carlo policy gradient algorithm}

  For each episode:
  \begin{enumerate}
  \item Generate a trajectory using $\pi_\theta$,
  \item Compute returns $G_t$,
  \item Update parameters:
    \[
    \theta \leftarrow \theta + \alpha
    \sum_t \nabla_\theta \log \pi_\theta(a_t|s_t)\,G_t
    \]
  \end{enumerate}

  Advantages:
  \begin{itemize}
  \item simple,
  \item direct,
  \item works with stochastic policies.
  \end{itemize}

  Disadvantage:
  \begin{itemize}
  \item large variance.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{Actor--Critic Methods}
  Actor--critic combines:
  \begin{itemize}
  \item \textbf{Actor}: policy $\pi_\theta(a|s)$,
  \item \textbf{Critic}: estimate of value function $V_\phi(s)$.
  \end{itemize}

  Temporal-difference error:
  \[
  \delta_t=r_{t+1}+\gamma V_\phi(s_{t+1})-V_\phi(s_t)
  \]

  Policy update:
  \[
  \theta \leftarrow \theta + \alpha_\theta
  \nabla_\theta \log \pi_\theta(a_t|s_t)\,\delta_t
  \]

  Critic update:
  \[
  \phi \leftarrow \phi - \alpha_\phi \nabla_\phi \delta_t^2
  \]

  This is often much more efficient than pure Monte Carlo policy gradients.
\end{frame}

%================================================
\begin{frame}{Worked RL Example: Grid World}
  Consider a particle moving on a 1D lattice:
  \[
  x \in \{0,1,2,3,4\}.
  \]

  Actions:
  \[
  a \in \{\text{left},\text{right}\}.
  \]

  Goal:
  \begin{itemize}
  \item reach $x=4$,
  \item avoid spending too many steps,
  \item possibly avoid a penalty region.
  \end{itemize}

  Reward example:
  \[
  r=
  \begin{cases}
    +10, & x=4,\\
    -1, & \text{each intermediate step}.
  \end{cases}
  \]

  This is simple enough to analyze explicitly, but already contains the core RL ideas.
\end{frame}

%================================================
\begin{frame}{Worked Example: Bellman Equations}
  If the policy is deterministic and always moves right, then
  \[
  V(4)=0
  \]
  if we stop at the goal, and for $x<4$,
  \[
  V(x)=-1+\gamma V(x+1).
  \]

  Thus
  \begin{align*}
    V(3) &= -1,\\
    V(2) &= -1+\gamma(-1),\\
    V(1) &= -1+\gamma(-1-\gamma),\\
    V(0) &= -1+\gamma(-1-\gamma-\gamma^2).
  \end{align*}

  So even this toy model already illustrates how future consequences are encoded recursively.
\end{frame}

%================================================
\begin{frame}{Worked Example: Q-Learning Equations}
  Suppose the agent experiences
  \[
  s_t=1,\qquad a_t=\text{right},\qquad r_{t+1}=-1,\qquad s_{t+1}=2.
  \]

  Then the update is
  \[
  Q(1,\text{right}) \leftarrow Q(1,\text{right}) +
  \alpha\left[-1 + \gamma \max_a Q(2,a)-Q(1,\text{right})\right].
  \]

  If initially all $Q=0$, then
  \[
  Q(1,\text{right}) \leftarrow \alpha(-1).
  \]

  Repeated experience gradually propagates information backward from the goal.
\end{frame}



%================================================
\begin{frame}[fragile]{Worked Example: Pseudocode}
  \small
\begin{verbatim}
  Initialize Q(s,a) arbitrarily
For each episode:
    Initialize state s
    While s is not terminal:
        Choose action a from Q using epsilon-greedy
        Take action a, observe reward r and next state s'
        Q(s,a) <- Q(s,a) + alpha *
                  [r + gamma * max_a' Q(s',a') - Q(s,a)]
        s <- s'
\end{verbatim}

\begin{itemize}
\item This is model-free.
\item No knowledge of transition probabilities is required.
\item The environment itself generates the data.
\end{itemize}
\end{frame}



%================================================
\begin{frame}[fragile]{Worked Example: Policy Gradient Pseudocode}
  \small
\begin{verbatim}
  Initialize policy parameters theta
  For each episode:
  Generate trajectory using pi_theta
  For each time step t:
  Compute return G_t
  Update:
          theta <- theta + alpha *
                 sum_t [grad_theta log pi_theta(a_t|s_t) * G_t]
\end{verbatim}

With a baseline:
\[
\theta \leftarrow \theta + \alpha
\sum_t \nabla_\theta \log \pi_\theta(a_t|s_t)\,
\big(G_t - b(s_t)\big)
\]

This becomes especially natural when actions are continuous.
\end{frame}




%================================================
\begin{frame}{Model-Free vs Model-Aware Reinforcement Learning}
  Two broad philosophies:

  \textbf{Model-free RL}
  \begin{itemize}
  \item learn directly from transitions,
  \item no explicit model of $P(s'|s,a)$ is used,
  \item examples: Q-learning, REINFORCE.
  \end{itemize}

  \textbf{Model-aware (or model-based) RL}
  \begin{itemize}
  \item use or learn a model of the dynamics,
  \item plan using that model,
  \item combine prediction and control.
  \end{itemize}

  In physics, model-aware RL is often especially attractive because we frequently know at least part of the dynamics.
\end{frame}

%================================================
\begin{frame}{What is Model-Aware Reinforcement Learning?}
  In model-aware RL we assume access to:
  \begin{itemize}
  \item known transition probabilities $P(s'|s,a)$, or
  \item an approximate learned model $\hat P_\phi(s'|s,a)$.
  \end{itemize}

  Then we can:
  \begin{itemize}
  \item simulate hypothetical futures,
  \item estimate returns before acting,
  \item plan over multiple time steps,
  \item use differentiable dynamics when available.
  \end{itemize}

  This is close in spirit to standard computational physics:
  \[
  \text{equations of motion} \;\Rightarrow\; \text{simulation} \;\Rightarrow\; \text{optimization}.
  \]
\end{frame}

%================================================
\begin{frame}{Benefits of Model-Aware RL}
  Advantages:
  \begin{itemize}
  \item better sample efficiency,
  \item ability to plan ahead,
  \item easier incorporation of prior physical knowledge,
  \item natural interface with control theory.
  \end{itemize}

  Disadvantages:
  \begin{itemize}
  \item model bias if the model is wrong,
  \item errors may accumulate during planning,
  \item learning an accurate model may be difficult.
  \end{itemize}

  Thus model-aware RL is especially useful when the dynamics are fairly well understood, but the control task remains hard.
\end{frame}

%================================================
\begin{frame}{Model-Aware RL and Physics}
  In many physics applications we know the dynamics:
  \[
  \dot x = f(x,u)
  \]
  or in quantum mechanics
  \[
  i\hbar \frac{d}{dt}\ket{\psi(t)}=H(u(t))\ket{\psi(t)}.
  \]

  Then the environment is not a black box:
  \begin{itemize}
  \item we can simulate trajectories,
  \item estimate gradients,
  \item evaluate candidate policies offline.
  \end{itemize}

  Model-aware RL thus sits naturally between:
  \begin{itemize}
  \item optimal control,
  \item stochastic optimization,
  \item machine learning.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{Transition to Quantum Systems}
  Quantum control problems are naturally sequential:
  \begin{itemize}
  \item apply pulse,
  \item observe state evolution,
  \item choose next pulse,
  \item optimize final performance.
  \end{itemize}

  This makes RL conceptually attractive.

  At the same time, quantum dynamics are often known from the Schr\"odinger equation, which makes model-aware RL particularly relevant.
\end{frame}

%================================================
\begin{frame}{Quantum Control}
  Quantum dynamics:
  \[
  i\hbar \frac{d}{dt}\ket{\psi(t)} = H(t)\ket{\psi(t)}
  \]

  Controlled Hamiltonian:
  \[
  H(t)=H_0+\sum_k u_k(t)H_k
  \]

  Typical goals:
  \begin{itemize}
  \item state preparation,
  \item gate synthesis,
  \item suppression of decoherence,
  \item enhancement of sensor performance.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{RL Formulation for Quantum Control}
  Mapping:
  \begin{itemize}
  \item \textbf{State}: wavefunction, density matrix, or measurement record,
  \item \textbf{Action}: control amplitude or pulse choice,
  \item \textbf{Reward}: fidelity, energy, Fisher information, robustness.
  \end{itemize}

  Examples of terminal rewards:
  \[
  R = |\langle \psi_{\mathrm{target}}|\psi(T)\rangle|^2
  \]
  or
  \[
  R = F_Q(T)
  \]
  for a quantum sensing task.
\end{frame}

%================================================
\begin{frame}{Quantum Sensors and Fisher Information}
  A quantum sensor estimates a parameter $\theta$.

  Precision is bounded by the quantum Cram\'er--Rao bound:
  \[
  \Delta \theta \ge \frac{1}{\sqrt{\nu F_Q}}
  \]
  where:
  \begin{itemize}
  \item $F_Q$ is the quantum Fisher information,
  \item $\nu$ is the number of repetitions.
  \end{itemize}

  Thus maximizing $F_Q$ is a natural control objective for reinforcement learning.
\end{frame}

%================================================
\begin{frame}{Worked Quantum Example: Two-Level Sensor}
  Consider a qubit sensing a magnetic field $B$:
  \[
  H(t)=\frac{\omega_0}{2}\sigma_z + \frac{B}{2}\sigma_z + u_x(t)\sigma_x.
  \]

  Interpretation:
  \begin{itemize}
  \item $\omega_0$: bare splitting,
  \item $B$: unknown signal,
  \item $u_x(t)$: control field chosen by the agent.
  \end{itemize}

  Goal:
  \begin{itemize}
  \item accumulate phase information about $B$,
  \item keep coherence as long as possible,
  \item maximize final sensitivity.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{Quantum Sensor as an RL Problem}
  Discretize time into steps $t_1,\dots,t_N$.

  Possible actions:
  \[
  a_t \in \{-u,0,+u\}.
  \]

  Possible state representation:
  \begin{itemize}
  \item Bloch vector $(\langle \sigma_x\rangle,\langle \sigma_y\rangle,\langle \sigma_z\rangle)$,
  \item time step,
  \item previous measurement outcomes.
  \end{itemize}

  Possible reward:
  \[
  R = F_Q(T)
  \]
  or
  \[
  R=\left|\frac{\partial}{\partial B}\langle \sigma_y(T)\rangle\right|.
  \]
\end{frame}

%================================================
\begin{frame}{Why RL for Quantum Sensors?}
  RL can learn how to:
  \begin{itemize}
  \item alternate between free precession and control,
  \item suppress unwanted evolution,
  \item balance phase accumulation against decoherence,
  \item adapt measurements based on partial information.
  \end{itemize}

  The learned strategy may resemble:
  \begin{itemize}
  \item Ramsey interferometry,
  \item spin echo,
  \item dynamical decoupling,
  \item adaptive phase estimation.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{Model-Aware RL for Quantum Control}
  Quantum systems are a prime example of model-aware RL.

  If the Hamiltonian is known, we can:
  \begin{itemize}
  \item simulate trajectories offline,
  \item evaluate hypothetical pulse sequences,
  \item use differentiable propagators,
  \item combine RL with standard optimal-control ideas.
  \end{itemize}

  This is especially attractive for:
  \begin{itemize}
  \item quantum control,
  \item calibration,
  \item sensor design,
  \item adaptive metrology.
  \end{itemize}
\end{frame}

%================================================
\begin{frame}{Summary}
  Reinforcement learning provides:
  \begin{itemize}
  \item a framework for sequential decision making,
  \item value-based and policy-based algorithms,
  \item a natural extension of stochastic optimization,
  \item direct relevance for quantum control and quantum sensing.
  \end{itemize}

  For physicists, the key conceptual bridge is:
  \[
  \text{variational optimization} + \text{stochastic dynamics} + \text{control}
  \;\Rightarrow\; \text{reinforcement learning}.
  \]
\end{frame}

%================================================


\end{document}



\begin{frame}{Appendix A: probe, controls, and measurement model}
A generic experiment is cast as
\[
  \rho \longmapsto \rho_{x,\theta}=\mathcal E_{x,\theta}(\rho),
\]
where $\rho$ is a reference probe state, $x$ denotes tunable controls, and $\theta\in\Theta$ are the unknown parameters.
\vspace{0.6em}
\begin{itemize}
  \item Controls can be continuous or discrete, and may represent durations, detunings, phases, or measurement settings.
  \item Parameters of interest can coexist with \emph{nuisance parameters}; estimating the latter can still improve the former.
  \item A measurement is modeled by a POVM $\mathcal M^x=\{M^x_y\}$ with likelihood
  \[
    p(y\mid \theta,x)=\mathrm{tr}\!\left(M^x_y\rho_{x,\theta}\right).
  \]
\end{itemize}
\end{frame}

\begin{frame}{Appendix A: projective versus weak measurements}
\begin{itemize}
  \item For \textbf{projective measurements}, the probe is reinitialized after each shot; the next likelihood depends only on $x_{t+1}$ and $\theta$.
  \item For \textbf{weak measurements}, information remains in the probe state, so the likelihood depends on the whole trajectory:
  \[
    p(y_{t+1}\mid x_{t+1},y_t,\theta)=\mathrm{tr}\!\left(M^{x_{t+1}}_{y_{t+1}}\,\rho_{x_t,y_t,\theta}\right).
  \]
  \item This turns the estimator into a partially observed control problem in which the posterior and probe state both carry memory.
\end{itemize}
\vspace{0.6em}
\textbf{Control lesson:} the state fed to the agent is not the raw experimental history, but a compressed belief-state summary derived from Bayesian inference.
\end{frame}

\begin{frame}{Appendix B: particle-filter representation of the posterior}
The posterior is approximated by a weighted ensemble
\[
  P(\theta)\approx\sum_{j=1}^N w_j^t\,\delta(\theta-\theta_j),
\]
with particles $\{\theta_j\}_{j=1}^N$ and weights $\{w_j^t\}_{j=1}^N$.
\vspace{0.4em}
\begin{itemize}
  \item Initialize particles from the prior $\pi(\theta)$ and set $w_j^0=1/N$.
  \item Apply Bayes updates after each observation:
  \[
    w_j^{t+1}=\frac{p(y_{t+1}\mid x_{t+1},\theta_j)w_j^t}{\sum_{i=1}^N p(y_{t+1}\mid x_{t+1},\theta_i)w_i^t}.
  \]
  \item The posterior mean and covariance become
  \[
    \hat\theta_t\approx\sum_j w_j^t\theta_j,
    \qquad
    \Sigma_t\approx\sum_j w_j^t(\theta_j-\hat\theta_t)(\theta_j-\hat\theta_t)^\top.
  \]
\end{itemize}
These moments form the default summary presented to the policy.
\end{frame}

\begin{frame}{Appendix B.3: resampling, soft resampling, and proposal moves}
\small
When weights collapse onto a few particles, the filter is refreshed by resampling.
\vspace{0.25em}
\begin{block}{Soft resampling}
Instead of sampling directly from $w_j$, define
\[
  q_j=\alpha w_j + (1-\alpha)\frac{1}{N},
\]
and draw new particles $\theta'_j=\theta_{\phi(j)}$ from $q_j$.
\end{block}
Corrected weights preserve the represented posterior:
\[
  w'_j\propto \frac{w_{\phi(j)}}{q_{\phi(j)}}.
\]
Further steps in Appendix B:
\begin{itemize}
  \item \textbf{Gaussian perturbation:} move particles toward $\hat\theta_t$ and break degeneracy.
  \item \textbf{New-particle proposal:} inject particles from $\mathcal N(\hat\theta_t,\Sigma_t)$ to raise local resolution.
\end{itemize}
\normalsize
\end{frame}

\begin{frame}{Appendix B.4: state particle filter for weak measurements}
For weak measurements, each parameter particle carries its own hypothetical probe state:
\[
  \rho_{\theta_j,\tau_t}=\mathcal M^{x_t}_{y_t}\circ\mathcal E_{\theta_j,x_t}\circ\cdots\circ
  \mathcal M^{x_1}_{y_1}\circ\mathcal E_{\theta_j,x_1}(\rho).
\]
The normalized measurement backaction map is
\[
  \mathcal M^x_y(\rho)=\frac{M^x_y\rho M^{x\dagger}_y}{\mathrm{tr}[M^x_y\rho M^{x\dagger}_y]}.
\]
\begin{itemize}
  \item The state particle filter stores $(\theta_j,w_j,\rho_{\theta_j,\tau_t})$.
  \item It enables trajectory-dependent likelihood evaluation for probes that are not reinitialized between measurements.
  \item This is the natural extension from Bayesian filtering over parameters to Bayesian filtering over \emph{parameter-conditioned states}.
\end{itemize}
\end{frame}

\begin{frame}{Why differentiation is hard}
The training objective depends on operations that are stochastic or discrete:
\begin{itemize}
  \item measurement outcomes $y_t\sim p(y_t\mid x_t,\theta)$,
  \item categorical particle resampling,
  \item stochastic perturbation and proposal draws,
  \item optional discrete decision outputs in classification tasks.
\end{itemize}
\vspace{0.5em}
Automatic differentiation treats sampled values as constants in the backward pass, so naively
\[
  \frac{d y_t}{d\lambda}=0,
\]
and the gradient would fail to capture how the policy changes the distribution of future trajectories.
\vspace{0.5em}
\textbf{Appendices C and D} supply the fix: combine score-function terms with differentiable surrogates wherever possible.
\end{frame}

\begin{frame}{Appendix C: differentiable particle filtering}
Three ingredients make the particle filter trainable.
\vspace{0.3em}
\begin{enumerate}
  \item \textbf{Soft resampling:} lets some gradient information survive resampling through the weights.
  \item \textbf{Reparametrization trick:} write perturbations and proposals as differentiable functions of Gaussian noise,
  \[
    \delta'(y_t,\lambda)=\Sigma(y_t,\lambda)u+\mu(y_t,\lambda),
    \qquad
    \theta''_j(y_t,\lambda)=\Sigma_t(y_t,\lambda)u_j+\hat\theta_t(y_t,\lambda).
  \]
  \item \textbf{Scibior-Wood correction:} replace the new weights by the surrogate
  \[
    \tilde w'_j \leftarrow w'_j\,\frac{q_{\phi(j)}}{\mathrm{sg}[q_{\phi(j)}]},
  \]
  which leaves the forward pass unchanged but injects missing backward terms.
\end{enumerate}
\vspace{0.4em}
For the MSE loss, the paper shows that this correction reproduces the needed score-function terms.
\end{frame}

\begin{frame}{Appendix D.1--D.3: Bayesian loss and policy-gradient correction}
For continuous estimation tasks, the single-run loss is an MSE-type quantity,
\[
  \ell(\hat\theta,\theta)=\mathrm{tr}\!\left[G(\hat\theta-\theta)(\hat\theta-\theta)^\top\right],
\]
while discrete tasks use a $0$-$1$ loss.
\vspace{0.4em}
Monte Carlo over a batch of $B$ simulations gives
\[
  L(\lambda)=\frac{1}{B}\sum_{k=1}^B \ell(\hat\theta_k,\theta_k).
\]
Because outcomes are sampled from a policy-dependent distribution, the correct training loss becomes
\[
  \tilde\ell(\omega_k,\lambda)=\ell(\omega_k,\lambda)
  +\mathrm{sg}[\ell(\omega_k,\lambda)]\log p(y_{k,0:M-1}\mid\theta_k,\lambda),
\]
so that the gradient includes both pathwise terms and score-function terms.
\end{frame}

\begin{frame}{Appendix D.4--D.6: cumulative and logarithmic losses}
A final-time loss alone may not yield good performance at intermediate resource budgets.
\vspace{0.5em}
\begin{block}{Cumulative loss}
\[
  L_{\mathrm{cum}}(\lambda)=\frac{1}{MB}\sum_{t=0}^{M-1}\sum_{k=1}^B
  \ell(\hat\theta_{k,t},\theta_k).
\]
This pressures the policy to use resources sensibly \emph{throughout} the experiment.
\end{block}
\begin{block}{Logarithmic loss}
Taking a logarithm of the batch-averaged loss stabilizes optimization when the precision spans several orders of magnitude.
\end{block}
\begin{itemize}
  \item A baseline can be subtracted inside the stop-gradient term to reduce gradient variance.
  \item The paper uses Adam rather than plain SGD for the actual parameter updates.
\end{itemize}
\end{frame}

\begin{frame}{Appendix E: simulation practice, hyperparameters, and scaling}
\textbf{Practical tuning strategy}
\begin{enumerate}
  \item Choose the number of particles $N$ large enough that posterior discretization is not the bottleneck.
  \item Use the remaining memory budget to set the batch size $B$.
  \item Tune the initial learning rate with the chosen loss (cumulative or logarithmic).
\end{enumerate}
\vspace{0.5em}
\textbf{Complexity trends reported in Appendix E.3}
\[
  \text{Memory}\sim \mathcal O(dN+N+n_\ell n_u),
  \qquad
  \text{Time}_{\mathrm{seq}}\sim \mathcal O(MN+Mn_\ell n_u^2),
\]
for parameter dimension $d$, measurements $M$, and a neural network with $n_\ell$ layers and $n_u$ units per layer.
\begin{itemize}
  \item Training benefits strongly from GPU parallelism over the batch.
  \item Deployment can be far cheaper, especially for non-adaptive strategies or offline Bayesian updates.
\end{itemize}
\end{frame}

\begin{frame}{Applications highlighted in the main text}
\textbf{NV-center DC magnetometry}
\begin{itemize}
  \item Control variable: Ramsey free-evolution time $\tau$.
  \item Likelihood: binary outcomes with dephasing-dependent contrast.
  \item Learned strategies beat common heuristics and prior model-free RL baselines in the reported settings.
\end{itemize}
\vspace{0.5em}
\textbf{Agnostic Dolinar receiver}
\begin{itemize}
  \item Task: discriminate coherent states while also treating the amplitude as an unknown nuisance parameter.
  \item Control variable: beam-splitter reflectivity/phase sequence.
  \item Adaptive RL approaches the finite-reference Helstrom limit more closely than prior hand-designed strategies.
\end{itemize}
\vspace{0.4em}
\textit{Figures intentionally omitted here; the corresponding plots from the article can be inserted later.}
\end{frame}

\begin{frame}{Appendix F: frequentist optimization via Fisher information}
The library also supports local frequentist design by optimizing the Cramer-Rao bound.
\vspace{0.4em}
\[
  K=\mathbb E\!\left[(\hat\theta-\theta)(\hat\theta-\theta)^\top\right]\ge F^{-1}(\theta),
\]
with trajectory-level Fisher information
\[
  F_{ij}(\theta)=\mathbb E_y\!\left[\frac{\partial\log p(y\mid x,\theta)}{\partial\theta_i}
  \frac{\partial\log p(y\mid x,\theta)}{\partial\theta_j}\right].
\]
Contracting with $G\ge 0$ gives the scalar loss
\[
  L(\lambda)=\mathrm{tr}\!\left(GF^{-1}\right).
\]
\begin{itemize}
  \item The gradient requires differentiating the Fisher matrix through the policy.
  \item Importance sampling can be used if rare trajectories dominate the Fisher information.
\end{itemize}
\end{frame}

\begin{frame}{Appendix G: lower bounds and how to benchmark learned policies}
Appendix G builds Bayesian Cramer-Rao style lower bounds tailored to the examples.
\vspace{0.4em}
\begin{itemize}
  \item For measurement-limited protocols,
  \[
    \Delta^2\hat\theta\ge \frac{1}{\sup_h \mathbb E_\theta[I(\theta\mid \tau_h)] + I(\pi)}
    \ge \frac{1}{M\,\mathbb E_\theta[\sup_\tau I(\theta\mid\tau)] + I(\pi)}.
  \]
  \item For time-limited protocols, the relevant object is the \emph{normalized} Fisher information $I(\theta\mid\tau)/\tau$.
  \item In NV magnetometry, binary outcomes also imply bit-counting limits that can dominate when the Fisher-information bound is loose.
\end{itemize}
\vspace{0.4em}
These bounds define the gray ``impossible region'' used in the paper's benchmark plots.
\end{frame}

\begin{frame}{Appendix H: backward recursion and the control-theory viewpoint}
Appendix H reframes the learned policy as an approximation to a dynamic-programming solution.
\vspace{0.4em}
\begin{itemize}
  \item At the last step, the optimal control $x^*_{M-1}$ minimizes the expected terminal loss conditioned on the current PF ensemble.
  \item Recursing backward defines a family of optimal decision rules $r_t(p_t)$ acting on the particle-filter state.
  \item The optimization is \textbf{non-myopic}: a control at time $t$ is chosen for its effect on the \emph{final} loss, not only the next measurement.
\end{itemize}
\vspace{0.5em}
\begin{block}{Interpretation}
The neural network is not a black box replacing physics; it is a function approximator for the intractable backward-recursion solution on belief states.
\end{block}
\end{frame}

\begin{frame}{Appendix I: backpropagation structure and the stop-gradient trick}
The terminal loss can be written as a composition of Bayes-update maps acting on the initial PF ensemble.
\vspace{0.5em}
The derivative expands into repeated products of two kinds of terms:
\begin{itemize}
  \item direct dependence of the Bayes update on the control selected by the agent,
  \item indirect dependence through earlier PF states fed back into later policy evaluations.
\end{itemize}
\vspace{0.5em}
Appendix I shows that:
\begin{itemize}
  \item these recursive terms generate expensive higher-order gradient paths,
  \item inserting a stop-gradient on the PF input,
  \[
    x_{t+1}=F_\lambda\big(\mathrm{sg}[P(\theta\mid x_t,y_t);R_t;t]\big),
  \]
  removes a large fraction of backward complexity,
  \item yet the resulting optimization remains non-myopic because each local update still targets final precision.
\end{itemize}
\end{frame}

\begin{frame}{Take-home messages}
\begin{enumerate}
  \item \textbf{Belief-state control:} quantum metrology can be formulated as sequential control on a Bayesian belief state.
  \item \textbf{Differentiable inference:} particle filtering can be embedded into model-aware RL through soft resampling, reparametrization, and score-function corrections.
  \item \textbf{Versatility:} the same framework supports adaptive/non-adaptive and Bayesian/frequentist design.
  \item \textbf{Interpretability:} Appendices H and I reveal the learned policy as an approximation to backward recursion with tractable backpropagation.
\end{enumerate}
\vspace{0.6em}
\textbf{Possible closing slide additions:}
benchmark plots, platform-specific diagrams, and an implementation workflow from \texttt{qsensoropt}.
\end{frame}



\end{document}
