\documentclass{beamer}
\usetheme{Madrid}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{listings}
\usepackage{xcolor}
\lstset{
   basicstyle=\ttfamily\footnotesize,
   backgroundcolor=\color{white},
   frame=single,
   breaklines=true
}
\title{Classifying Phase Transitions with Machine Learning}
\author{Morten Hjorth-Jensen}
\date{\today}

\begin{document}

\begin{frame}
 \titlepage
\end{frame}

\begin{frame}{Outline}
 \tableofcontents
\end{frame}

\section{Phase Transitions and Critical Phenomena}

\begin{frame}{Phase Transitions: Definitions}
 \begin{itemize}
   \item A \emph{phase transition} is a qualitative change in the state of a system when a control parameter (e.g.\ temperature) passes a critical point.
   \item \textbf{Order parameter}: quantity that distinguishes phases (e.g.\ magnetization $M$ for magnetic systems).
   \item \textbf{Order vs disorder}: e.g.\ below $T_c$ a ferromagnet has $|M|>0$ (ordered), above $T_c$ $M=0$ (disordered).
   \item Phases can break symmetries; transitions can be \emph{continuous} (second-order) or \emph{first-order}.
 \end{itemize}
\end{frame}

\begin{frame}{Order Parameter \& Symmetry Breaking}
 \begin{itemize}
   \item Phase transitions often involve spontaneous symmetry breaking (e.g.\ Ising model $Z_2$ symmetry).
   \item The order parameter (e.g.\ magnetization $M = \frac{1}{N}\sum_i s_i$) changes behavior at $T_c$.
   \item In ferromagnets: $M=0$ for $T>T_c$ (symmetric paramagnet), $M\neq 0$ for $T<T_c$ (broken symmetry).
   \item Example: in 2D Ising model, two symmetric ordered states (up/down) below $T_c$.
 \end{itemize}
\end{frame}

\begin{frame}{Critical Phenomena and Scaling}
 \begin{itemize}
   \item Near a continuous transition, observables follow power laws: $M \sim |T-T_c|^\beta$, correlation length $\xi \sim |T-T_c|^{-\nu}$, etc.
   \item \textbf{Critical exponents} ($\alpha,\beta,\gamma,\nu,\dots$) characterize singular behavior.
   \item Universality: systems with the same symmetry and dimension share exponents.
   \item E.g.\ 2D Ising exponents known analytically (Onsager).
   \item At $T\to T_c$, correlation length $\xi\to\infty$, large-scale fluctuations appear.
 \end{itemize}
\end{frame}

\section{Spin Models: 2D Ising and Potts}

\begin{frame}{2D Ising Model: Definition}
 \begin{itemize}
   \item Spins $s_i = \pm 1$ on a 2D square lattice, nearest-neighbor ferromagnetic coupling.
   \item Hamiltonian:
     \[
       H = -J \sum_{\langle i,j\rangle} s_i s_j,
     \]
     with $J>0$ favoring alignment.
   \item Exhibits a second-order phase transition at critical temperature $T_c$.
   \item Order parameter: magnetization $M = \frac{1}{N}\sum_i s_i$.
   \item Below $T_c$, $M\neq0$ (ferromagnetic order); above $T_c$, $M=0$ (paramagnet).
 \end{itemize}
\end{frame}

\begin{frame}{2D Ising Model: Critical Temperature}
 \begin{itemize}
   \item Exact result (Onsager): critical point $T_c$ satisfies
     \[
       \frac{2J}{k_B T_c} = \ln(1+\sqrt{2}) \quad\Rightarrow\quad T_c \approx \frac{2J}{\ln(1+\sqrt{2})}\approx 2.269J.
     \]
     (For simplicity set $k_B=1$) [oai_citation:0‡sites.pitt.edu](https://sites.pitt.edu/~jdnorton/teaching/philphys/Ising_sim/index.html#:~:text=,is%20set%20to%20this%20value).
   \item At $T>T_c$: spins are mostly disordered, no net magnetization.
   \item At $T<T_c$: long-range order develops (nonzero $M$).
   \item Correlation length $\xi$ diverges at $T_c$ [oai_citation:1‡sites.pitt.edu](https://sites.pitt.edu/~jdnorton/teaching/philphys/Ising_sim/index.html#:~:text=4,diverges%20at%20the%20critical%20point).
   \item Example: at $T=T_c$ large clusters of aligned spins appear.
 \end{itemize}
\end{frame}

\begin{frame}{q-State Potts Model: Definition}
 \begin{itemize}
   \item Generalization of Ising: each spin $s_i \in \{1,2,\dots,q\}$.
   \item Ferromagnetic Potts Hamiltonian:
     \[
       H = -J \sum_{\langle i,j\rangle} \delta_{s_i,s_j},
     \]
     where $\delta_{a,b}=1$ if $a=b$, else $0$.
   \item If $q=2$, reduces to the Ising model. Higher $q$ allows richer symmetry breaking ($\mathbb{Z}_q$).
   \item Widely used to study phase transitions with multiple equivalent ordered states.
 \end{itemize}
\end{frame}

\begin{frame}{2D Potts Model: Phase Behavior}
 \begin{itemize}
   \item In 2D, the ferromagnetic Potts model has a phase transition for all $q\ge1$ [oai_citation:2‡en.wikipedia.org](https://en.wikipedia.org/wiki/Potts_model#:~:text=for%20the%20standard%20ferromagnetic%20Potts,4%7D.%5B%209).
   \item Exact critical point:
     \[
       \frac{J}{k_B T_c} = \ln\!\bigl(1+\sqrt{q}\bigr).
     \]
   \item The nature of the transition depends on $q$ [oai_citation:3‡en.wikipedia.org](https://en.wikipedia.org/wiki/Potts_model#:~:text=for%20the%20standard%20ferromagnetic%20Potts,4%7D.%5B%209):
     \begin{itemize}
       \item $1 \le q \le 4$: continuous (second-order) transition.
       \item $q > 4$: discontinuous (first-order) transition (latent heat appears).
     \end{itemize}
   \item Example: $q=3,4$ have continuous transitions; $q=5$ and higher show first-order behavior.
 \end{itemize}
\end{frame}

\section{Monte Carlo Simulation}

\begin{frame}{Monte Carlo Sampling of Spin Models}
 \begin{itemize}
   \item Use Monte Carlo (MC) to generate spin configurations at given $T$:
     sample from Boltzmann distribution $P\propto e^{-H/T}$.
   \item Metropolis algorithm: attempt random single-spin flips to equilibrate the system.
   \item Provides training data: spin configurations $\{s_i\}$ labeled by temperature or phase.
   \item Ensures statistical equilibrium and detailed balance [oai_citation:4‡rajeshrinet.github.io](https://rajeshrinet.github.io/blog/2014/ising-model/#:~:text=1,4).
   \item Efficient sampling (especially near $T_c$ cluster algorithms help, e.g.\ Wolff or Swendsen-Wang).
 \end{itemize}
\end{frame}

\begin{frame}{Metropolis Algorithm (Concept)}
 \begin{itemize}
   \item Initialize spins randomly or in a fixed state.
   \item Repeat for many steps:
     \begin{enumerate}
       \item Pick a random lattice site $i$.
       \item Propose flipping $s_i \to -s_i$ (Ising) or change state (Potts).
       \item Compute energy change $\Delta E$.
       \item If $\Delta E \le 0$, accept the flip (lower energy).
       \item Else accept with probability $\exp(-\Delta E/T)$ (Boltzmann factor) [oai_citation:5‡rajeshrinet.github.io](https://rajeshrinet.github.io/blog/2014/ising-model/#:~:text=1,4).
       \item Otherwise, reject and keep the old state.
     \end{enumerate}
   \item After equilibration, record configurations as samples.
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{Metropolis Algorithm (Pseudo-code)}
 \begin{lstlisting}[language=Python]
for T in temperature_list:
   # Initialize lattice (e.g., random spins)
   config = random_configuration(Lx, Ly)
   for step in range(num_steps):
       i,j = random_site()
       dE = compute_deltaE(config, i, j)  # energy change if spin flipped
       if dE <= 0 or rand() < exp(-dE/T):
           flip_spin(config, i, j)
   record_configuration(config, T)
 \end{lstlisting}
\end{frame}

\begin{frame}{Monte Carlo Data for ML}
 \begin{itemize}
   \item Generate many spin configurations across a range of temperatures $T$.
   \item Label each configuration by its temperature or by phase (ordered/disordered).
   \item This labeled dataset is used for \emph{supervised} methods (e.g.\ CNN).
   \item For \emph{unsupervised} methods (PCA, VAE), labels are not used in training.
   \item Data augmentation: one can use symmetries (e.g.\ spin flip) to enlarge dataset.
 \end{itemize}
\end{frame}

\section{Unsupervised Learning: PCA}

\begin{frame}{Principal Component Analysis (PCA) Basics}
 \begin{itemize}
   \item PCA is an unsupervised method for dimensionality reduction.
   \item Finds orthogonal directions (principal components) of maximum variance in data.
   \item Project data onto the first few PCs to visualize structure.
   \item Advantages: linear, fast, and interpretable (PCs are linear combinations of features).
   \item Disadvantage: only captures linear correlations (may miss complex features).
 \end{itemize}
\end{frame}

\begin{frame}{PCA for Phase Identification}
 \begin{itemize}
   \item Apply PCA to the ensemble of spin configurations (flattened to vectors).
   \item The first principal component (PC1) often correlates with the order parameter (e.g.\ magnetization).
   \item Hu et al. (2017) found PCA distinguishes different phases and can locate critical points [oai_citation:6‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
   \item By plotting data in the subspace of PCs, one sees separation of low-$T$ (ordered) vs high-$T$ (disordered) points.
   \item No labels needed: phase transitions are revealed by clustering in PC space [oai_citation:7‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
 \end{itemize}
\end{frame}

\begin{frame}{PCA Workflow for Spin Data}
 \begin{itemize}
   \item Collect data matrix $X$ of shape (num\_samples) $\times$ (num\_features), e.g. $N\times (L\times L)$.
   \item Subtract the mean from each column (feature) of $X$.
   \item Compute covariance matrix $C = X^T X$ (or use SVD on $X$ directly).
   \item Obtain eigenvalues/vectors of $C$: $C = U \Lambda U^T$. Columns of $U$ are principal directions.
   \item Sort by eigenvalues (variance). Project $X$ onto top $k$ PCs: $X_{\rm red} = X\,U[:,1:k]$.
   \item Analyze $X_{\rm red}$: e.g.\ scatter plot PC1 vs PC2.
 \end{itemize}
\end{frame}

\begin{frame}{PCA Example: Ising Model}
 \begin{itemize}
   \item In the 2D Ising model, PC1 is essentially proportional to the overall magnetization.
   \item At $T < T_c$, configurations cluster with large positive or negative PC1 (ordered states).
   \item At $T > T_c$, configurations cluster near PC1 ≈ 0 (disordered).
   \item The variance captured by PC1 drops sharply at $T_c$, signaling the transition.
   \item PCA automatically finds these features, without knowing the physics a priori [oai_citation:8‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
 \end{itemize}
\end{frame}

\begin{frame}{PCA Limitations}
 \begin{itemize}
   \item PCA is linear: complex nonlinear features (e.g.\ vortex order) may not be captured.
   \item Example: In a frustrated 2D spin model, PCA failed to detect certain correlations (vorticity) [oai_citation:9‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=well%20known,phase%20transitions%20and%20critical%20points).
   \item PCA does not directly classify; it provides features for clustering or visualization.
   \item Sensitive to scaling: data should be normalized appropriately.
   \item Still useful as a first-pass: identifies the most significant variations [oai_citation:10‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{PCA with PyTorch (Example Code)}
 \begin{lstlisting}[language=Python]
import torch

# X: tensor of shape (N, L*L) containing spin configurations as floats (e.g. +1/-1)
# Center the data
X = X - X.mean(dim=0, keepdim=True)

# Compute covariance (or use torch.pca_lowrank)
cov = torch.mm(X.t(), X) / (X.size(0)-1)

# Eigen-decomposition (SVD) of covariance
U, S, V = torch.svd(cov)

# Select first k principal components
k = 2
PCs = U[:, :k]  # shape (L*L, k)

# Project data onto principal components
X_reduced = torch.mm(X, PCs)  # shape (N, k)
 \end{lstlisting}
\end{frame}

\section{Supervised Learning: CNNs}

\begin{frame}{Convolutional Neural Networks (CNNs)}
 \begin{itemize}
   \item CNNs are deep neural networks designed for spatial data (e.g.\ images).
   \item Architecture: convolutional layers (feature detectors) + pooling, followed by fully connected layers.
   \item In physics: treat spin lattice as an image with multiple channels (e.g.\ one channel of spins).
   \item CNNs can learn complex nonlinear features automatically from data.
   \item They require labeled examples for training (supervised learning).
 \end{itemize}
\end{frame}

\begin{frame}{CNN for Phase Classification}
 \begin{itemize}
   \item Prepare training data: spin configurations labeled by phase or temperature.
   \item CNN learns to map configuration $\to$ phase label (ordered/disordered) or predict $T$.
   \item As shown by Carrasquilla and Melko (2017), CNNs can identify phases from raw states [oai_citation:11‡nature.com](https://www.nature.com/articles/nphys4035#:~:text=abilities%20to%20recognize%2C%20classify%2C%20and,sampled%20with%20Monte%20Carlo%2023%2C7).
   \item Achieves high accuracy on Ising and other models when training labels are available.
   \item CNNs exploit locality: can detect clusters or domains of aligned spins via convolution filters.
 \end{itemize}
\end{frame}

\begin{frame}{Example CNN Architecture}
 \begin{itemize}
   \item \textbf{Input}: single-channel $L\times L$ lattice (values $-1$ or $+1$).
   \item \textbf{Conv layer 1}: e.g.\ 8 filters of size $3\times3$, ReLU activation, stride=1, padding=1.
   \item \textbf{Conv layer 2}: 16 filters of size $3\times3$, ReLU, followed by a $2\times2$ max-pooling.
   \item \textbf{Fully Connected}: flatten feature maps to vector; FC layer to 64 units (ReLU); final FC to 2 outputs (softmax for binary phase).
   \item \textbf{Training}: minimize cross-entropy loss between predicted and true labels.
   \item \textbf{Note}: architecture and hyperparameters can be tuned for best performance.
 \end{itemize}
\end{frame}

\begin{frame}{CNN: Training and Results}
 \begin{itemize}
   \item Train on many labeled samples (e.g.\ temperatures $T$ and whether $T<T_c$ or $T>T_c$).
   \item The network learns features such as magnetization domains, energy patterns, etc.
   \item CNN accuracy can be very high (often $\sim$100\% on clean data) for distinguishing phases.
   \item Fukushima \& Sakai (2021): a CNN trained on 2D Ising can detect transition in $q$-state Potts [oai_citation:12‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity).
   \item CNN behavior: at high $T$ it effectively uses average energy; at low $T$ it correlates with magnetization [oai_citation:13‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=accuracy%2C%20regardless%20of%20the%20type,However%2C%20in%20the%20vicinity).
 \end{itemize}
\end{frame}

\begin{frame}{CNN Interpretability}
 \begin{itemize}
   \item CNNs are often seen as "black boxes", but their learned filters can sometimes be interpreted.
   \item Outputs correlate with known physics:
     \begin{itemize}
       \item At low $T$: classification heavily influenced by magnetization (order).
       \item At high $T$: classification influenced by internal energy (disorder) [oai_citation:14‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=accuracy%2C%20regardless%20of%20the%20type,However%2C%20in%20the%20vicinity).
     \end{itemize}
   \item CNNs can generalize: e.g.\ Ising-trained CNN finds Potts $T_c$ [oai_citation:15‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity), implying it learned universal features.
   \item Visualization methods (e.g.\ saliency maps) can highlight what CNN focuses on.
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{CNN (PyTorch) Code Example}
 \begin{lstlisting}[language=Python]
import torch
import torch.nn as nn
import torch.nn.functional as F

class PhaseCNN(nn.Module):
   def __init__(self, L):
       super(PhaseCNN, self).__init__()
       self.conv1 = nn.Conv2d(1, 8, kernel_size=3, padding=1)  # 1 channel -> 8
       self.conv2 = nn.Conv2d(8, 16, kernel_size=3, padding=1) # 8 -> 16
       self.pool = nn.MaxPool2d(2)  # downsample by 2
       self.fc1 = nn.Linear(16 * (L//2) * (L//2), 64)
       self.fc2 = nn.Linear(64, 2)  # 2 output classes

   def forward(self, x):
       x = F.relu(self.conv1(x))      # (B,8,L,L)
       x = self.pool(F.relu(self.conv2(x)))  # (B,16,L/2,L/2)
       x = x.view(x.size(0), -1)      # flatten
       x = F.relu(self.fc1(x))
       x = self.fc2(x)               # logits for 2 classes
       return x

# Example usage:
model = PhaseCNN(L=32)           # for a 32x32 lattice
criterion = nn.CrossEntropyLoss()
optimizer = torch.optim.Adam(model.parameters(), lr=1e-3)
 \end{lstlisting}
\end{frame}

\section{Unsupervised Learning: Variational Autoencoders}

\begin{frame}{Variational Autoencoders (VAE) Overview}
 \begin{itemize}
   \item A VAE is an \emph{unsupervised} generative model that learns a latent representation of data.
   \item Components:
     \begin{itemize}
       \item \textbf{Encoder}: maps input $X$ to parameters $(\mu, \log\sigma^2)$ of a latent Gaussian.
       \item \textbf{Latent $z$}: sampled via $z = \mu + \sigma \epsilon$ ($\epsilon\sim N(0,I)$).
       \item \textbf{Decoder}: reconstructs input $\hat{X}$ from $z$.
     \end{itemize}
   \item Loss: reconstruction error + KL divergence to enforce latent prior $\mathcal{N}(0,I)$.
   \item VAEs can both encode data compactly and generate new samples by sampling $z$.
 \end{itemize}
\end{frame}

\begin{frame}{VAE for Spin Configurations}
 \begin{itemize}
   \item Train VAE on spin configurations (no labels).
   \item Latent space (usually low-dimensional) captures key features (like order parameter).
   \item Walker et al. (2020): latent variables provide metrics to track order vs disorder in Ising [oai_citation:16‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item They found the latent representation closely corresponds to physical order (magnetization) [oai_citation:17‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations).
   \item After training, one can:
     \begin{itemize}
       \item Inspect latent space (e.g.\ scatter plot of $(\mu_1,\mu_2)$) to distinguish phases.
       \item Sample $z\sim N(0,1)$ and decode to generate synthetic configurations.
     \end{itemize}
 \end{itemize}
\end{frame}

\begin{frame}{VAE Architecture Details}
 \begin{itemize}
   \item Typically use convolutional encoder/decoder for 2D structure.
   \item Example:
     \begin{itemize}
       \item Encoder: conv layers downsampling to a flat vector $\rightarrow$ linear layers $\rightarrow (\mu, \log\sigma^2)$ (size of latent space, e.g.\ 2–10 dims).
       \item Decoder: linear layer from $z$ to feature map size, followed by transposed-conv layers to reconstruct $L\times L$ lattice.
     \end{itemize}
   \item Activation: ReLU (or LeakyReLU); final output often sigmoid to model spin distribution.
   \item Training with minibatch gradient descent optimizing
     \[
       \mathcal{L} = \mathbb{E}[\|X - \hat{X}\|^2] +
       \mathrm{KL}(\mathcal{N}(\mu,\sigma)\,\|\,\mathcal{N}(0,1)).
     \]
 \end{itemize}
\end{frame}

\begin{frame}{VAE Results on Ising Model}
 \begin{itemize}
   \item The first latent dimension ($\nu_0$) learned by the VAE correlated strongly with magnetization [oai_citation:18‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations).
   \item Plotting $\nu_0$ vs temperature shows clear change around $T_c$ (order–disorder).
   \item This means VAE "discovered" the order parameter without supervision.
   \item The VAE predicted the critical region and crossover consistently with theory [oai_citation:19‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item Latent space clustering: ordered-phase points separate from disordered.
 \end{itemize}
\end{frame}

\begin{frame}{VAE: Generation and Interpretation}
 \begin{itemize}
   \item After training, sample random $z$ from Gaussian prior and decode to generate configurations.
   \item The VAE latent space is continuous: can interpolate between phases.
   \item The learned representation is smooth and disentangled: one latent coordinate tracks magnetization, others track disorder.
   \item VAEs can also be used for anomaly detection: points with unusual $z$ indicate atypical states.
   \item Overall, VAEs provide both a dimensionally-reduced view of phase structure and a generative model.
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{VAE (PyTorch) Code Example}
 \begin{lstlisting}[language=Python]
import torch
import torch.nn as nn
import torch.nn.functional as F

class VAE(nn.Module):
   def __init__(self, L, latent_dim=2):
       super(VAE, self).__init__()
       # Encoder: conv -> conv -> flatten -> fc_mu/fc_logvar
       self.encoder = nn.Sequential(
           nn.Conv2d(1, 8, 3, stride=2, padding=1),   # -> (8, L/2, L/2)
           nn.ReLU(),
           nn.Conv2d(8, 16, 3, stride=2, padding=1),  # -> (16, L/4, L/4)
           nn.ReLU(),
           nn.Flatten()
       )
       self.fc_mu = nn.Linear(16*(L//4)*(L//4), latent_dim)
       self.fc_logvar = nn.Linear(16*(L//4)*(L//4), latent_dim)

       # Decoder: linear -> unflatten -> convTranspose -> convTranspose
       self.decoder_fc = nn.Linear(latent_dim, 16*(L//4)*(L//4))
       self.decoder = nn.Sequential(
           nn.Unflatten(1, (16, L//4, L//4)),
           nn.ConvTranspose2d(16, 8, 3, stride=2, padding=1, output_padding=1),  # -> (8, L/2, L/2)
           nn.ReLU(),
           nn.ConvTranspose2d(8, 1, 3, stride=2, padding=1, output_padding=1),   # -> (1, L, L)
           nn.Sigmoid()
       )

   def encode(self, x):
       h = self.encoder(x)
       mu = self.fc_mu(h)
       logvar = self.fc_logvar(h)
       return mu, logvar

   def reparameterize(self, mu, logvar):
       std = torch.exp(0.5*logvar)
       eps = torch.randn_like(std)
       return mu + eps * std

   def decode(self, z):
       h = self.decoder_fc(z)
       x_recon = self.decoder(h)
       return x_recon

   def forward(self, x):
       mu, logvar = self.encode(x)
       z = self.reparameterize(mu, logvar)
       x_recon = self.decode(z)
       return x_recon, mu, logvar

# Example instantiation:
model = VAE(L=32, latent_dim=2)
optimizer = torch.optim.Adam(model.parameters(), lr=1e-3)
 \end{lstlisting}
\end{frame}

\section{Comparison of Methods}

\begin{frame}{Supervised vs Unsupervised Methods}
 \begin{itemize}
   \item \textbf{Supervised (CNN)}: Requires labeled data (phase labels or temperatures). Learns a direct mapping $\{\text{config}\}\to\{\text{phase}\}$.
   \item \textbf{Unsupervised (PCA, VAE)}: Uses only the raw configurations without labels. Learns features or representations of the data.
   \item PCA reduces dimensionality; requires no training labels [oai_citation:20‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
   \item VAE learns a latent generative model; also label-free [oai_citation:21‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item CNN typically achieves higher accuracy in classifying known phases, but needs supervised labels.
 \end{itemize}
\end{frame}

\begin{frame}{Method Interpretability and Features}
 \begin{itemize}
   \item \textbf{PCA}: Principal components often have clear physical meaning (e.g.\ PC1 ∝ magnetization) [oai_citation:22‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
   \item \textbf{CNN}: Filters are less directly interpretable; features are learned. However, some correlation with physics (energy, $M$) emerges [oai_citation:23‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=accuracy%2C%20regardless%20of%20the%20type,However%2C%20in%20the%20vicinity).
   \item \textbf{VAE}: Latent variables can often be interpreted as order/disorder features (e.g.\ one latent ≈ magnetization) [oai_citation:24‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations).
   \item CNN is a "black box" classifier; PCA/VAE provide insight into data structure.
   \item In terms of visualization: PCA and VAE produce low-dim plots of data (semi-transparent), whereas CNN only outputs a decision boundary.
 \end{itemize}
\end{frame}

\begin{frame}{Performance and Use Cases}
 \begin{itemize}
   \item \textbf{PCA}: Fast to compute; good for preliminary analysis of large datasets. Best for linearizable transitions.
   \item \textbf{CNN}: High classification accuracy; powerful for large and complex datasets. Can predict critical $T$ or classify multiple phases [oai_citation:25‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity).
   \item \textbf{VAE}: Useful when no labels are available; provides a generative model. Effective in detecting transitions by latent statistics [oai_citation:26‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item Computational cost: PCA very cheap, CNN and VAE require training time (GPU recommended for large data).
   \item Choosing a method: depends on data availability and goal (classification vs insight vs generation).
 \end{itemize}
\end{frame}

\begin{frame}{Example Case Studies}
 \begin{itemize}
   \item \textbf{Ising/CNN}: CNNs trained on Ising achieve $\approx100\%$ phase classification accuracy.
   \item \textbf{Ising $\to$ Potts}: Fukushima \& Sakai show an Ising-trained CNN still identifies $T_c$ in Potts models [oai_citation:27‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity).
   \item \textbf{PCA on Potts}: PCA can also locate $T_c$ for Potts by looking at variance of PCs.
   \item \textbf{VAE on Ising}: Walker et al.\ VAE found latent coordinate correlates with magnetization; predicted crossover region [oai_citation:28‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations) [oai_citation:29‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item \textbf{Limitation example}: PCA failed to find transition in a frustrated spin model due to non-linear order [oai_citation:30‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=well%20known,phase%20transitions%20and%20critical%20points).
 \end{itemize}
\end{frame}

\begin{frame}{Summary of Methods}
 \begin{itemize}
   \item \textbf{PCA}: Unsupervised, linear, interpretable. Good for dimensionality reduction and initial exploration [oai_citation:31‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
   \item \textbf{CNN}: Supervised, non-linear, high accuracy. Requires labels, but learns complex features (works across models [oai_citation:32‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity)).
   \item \textbf{VAE}: Unsupervised, generative. Learns latent representation reflecting order/disorder [oai_citation:33‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations). Good for unsupervised detection and data generation.
   \item Each method has trade-offs in accuracy, interpretability, and data requirements.
   \item Combining methods (e.g.\ using PCA or VAE features as input to another classifier) can also be fruitful.
 \end{itemize}
\end{frame}

\section{Conclusions}

\begin{frame}{Conclusions}
 \begin{itemize}
   \item Machine learning provides powerful tools for studying phase transitions in statistical models.
   \item \emph{Unsupervised} methods (PCA, VAE) can discover phase structure without labels [oai_citation:34‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical) [oai_citation:35‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method).
   \item \emph{Supervised} methods (CNNs) achieve high classification performance given labeled data [oai_citation:36‡nature.com](https://www.nature.com/articles/nphys4035#:~:text=abilities%20to%20recognize%2C%20classify%2C%20and,sampled%20with%20Monte%20Carlo%2023%2C7).
   \item Interpretability: PCA/VAE offer more insight into physics (latent/PC represent order parameters), while CNNs focus on prediction accuracy.
   \item Choice of method depends on the problem: data availability, need for generative modeling, and interpretability.
   \item Future directions: deeper architectures (e.g.\ ResNets), unsupervised generative flows, transfer learning across models, real experimental data.
 \end{itemize}
\end{frame}

\begin{frame}{References}
 \begin{itemize}
   \item Carrasquilla, J. \& Melko, R. G. (2017). Machine learning phases of matter. \emph{Nature Physics}, 13, 431–434 [oai_citation:37‡nature.com](https://www.nature.com/articles/nphys4035#:~:text=abilities%20to%20recognize%2C%20classify%2C%20and,sampled%20with%20Monte%20Carlo%2023%2C7).
   \item Hu, W. \textit{et al.} (2017). Discovering phases, phase transitions through unsupervised ML. \emph{Phys. Rev. E} 95, 062122 [oai_citation:38‡link.aps.org](https://link.aps.org/doi/10.1103/PhysRevE.95.062122#:~:text=We%20apply%20unsupervised%20machine%20learning,vectors%20have%20a%20clear%20physical).
   \item Fukushima, K. \& Sakai, K. (2021). Can a CNN trained on Ising detect Potts? \emph{Prog. Theor. Exp. Phys.} 2021, 061A01 [oai_citation:39‡arxiv.org](https://arxiv.org/abs/2104.03632#:~:text=generated,However%2C%20in%20the%20vicinity).
   \item Walker, N. \textit{et al.} (2020). 2D Ising model crossover via VAE. \emph{Sci. Rep.} 10, 13047 [oai_citation:40‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=The%202,The%20performance%20of%20this%20method) [oai_citation:41‡nature.com](https://www.nature.com/articles/s41598-020-69848-5#:~:text=By%20comparing%20,inaccuracies%20in%20the%20intermediate%20magnetizations).
   \item [Additional references on PCA, CNN, VAE methods].
 \end{itemize}
\end{frame}

\end{document}
